---
title: "Prior Choice"
subtitle: "how to choose a good prior"
author: "Shangchen Song"
institute: "University of Florida"
date: "2020/12/01 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
library(webshot)
style_mono_accent(
  base_color = "#1c5253",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Fira Mono")
)
```


```{r}

knitr::include_url("https://baidu.com")
```

### What is Prior Distribution


The prior distribution is a key part of Bayesian inference and represents the information about an uncertain parameter $\theta$ that is combined with the probability distribution of new data to yield the **posterior distribution**, which in turn is used for future inferences and decisions involving $\theta$.


<br/>
$$
\text{posterior} \propto  \text{likelihood} \times \text{prior}
$$

---



### Why the Prior Distribution is important

- Unreasonable prior may cause improper posterior distribution.

- For all $\theta$ for which $\pi(\theta) = 0$, we necessarily have $p(\theta | y) = 0$. If the prior does not contain all range of the parameter, the posterior will not consider the excluding part.

- If the sample size is small, the posterior distribution will utilize the information from the prior distribution considerably.
    - although the larger the sample size is, the smaller the effect is. 


???

The next is the reason why the prior distribution is important?

- an important first observation is that For all $\theta$ for which the prior pdf, $\pi(\theta) = 0$, we necessarily have the posterior pdf, $p(\theta | y), = 0$. Which means that, if we previous believe that the parameter is not possible to appear than it would not appear in the posterior distribution.


- second, If the sample size is small, the prior distribution is very important. So, we have to be careful to specify a reasonable prior. One remedy methods is that models can be set up hierarchically, so that clusters of parameters have shared prior distributions, which can themselves be estimated from previous data.

- With large sample sizes, reasonable choices of prior distributions will have minor effects on posterior inferences. In practice one can check the dependence on prior distributions by a sensitivity analysis: comparing posterior inferences under different reasonable choices of prior distribution

---

### Issue to Think for the Prior Choice


1. What information is going into the prior distribution.

  - must include all possible values of the parameter of interest.
  - should include reasonable expert information we have for the parameter.
<br/><br/>
2. The properties of the resulting posterior distribution.

  - whether the posterior is proper or not proper;
  - whether the posterior is closed-form or not, although this can be overcome by MCMC algorithm.
  
---

class: inverse, middle, center

# The Types of The Prior Distribution

---
### Noninformative Prior

- It is also called as **diffuse/default/low-informative/uniform/objective** prior 

- it essentially reflects a lack of strong and precisely quantified prior information.

- it assigns a relatively equal likelihood among outcomes. It brings minimal impact for the posterior distribution so that the information in the likelihood dominates the posterior.

- the resulting posterior distribution can have good frequentist properties.

- it can take some principles like such as symmetry or maximizing entropy given constraints; examples are the Jeffreys prior or Bernardo's reference prior. 

- when a family of conjugate priors exists, choosing a prior from that family simplifies calculation of the posterior distribution. (Super-vague but proper prior)

- it may cause unexpected parameter value. (like tomorrow's temperature could be 100 degree with small probability)
---

### Noninformative Prior: Flat Prior

A **flat prior** is a uniform prior on the parameter: $f(θ)$ is constant or written as $f(θ) \propto 1$



  - straightforward

  - the final posterior could be improper.
    - e.g. for the nonlinear models, the flat prior should never be used, because it will often result in the improper posterior.



---

### Example
Suppose that $Y_1, . . . , Y_n$ are from  ** i.i.d.** $N(\mu, 1)$, we use a “prior” $f(\mu) = 1$.


Of course this prior is not a probability distribution, but it turns
out that we can still formally carry out the calculation of a
posterior, as follows:

- The product of the likelihood and the prior is just the
likelihood.

- If we can normalize the likelihood to be a probability distribution over $\mu$, then this will be a well-defined posterior.

Note that in this case the posterior is just a scaled version of likelihood, so the **posterior mode** is exactly the same as the **MLE**

---

### Noninformative Prior: Jeffery's Prior

The format is $$\pi(\theta) \propto |I(\theta)|^{1/2}$$ where $I(\theta)$ is Fisher's expected information.


- it has the desirable property of invariance of reparameterization, regardless of the univaraite or multivariate setting.

- however, it may lead to the posterior with undesirable characteristics in multivariate case.
  - like the Neyman-Scott problem in our homework, the use of Jeffery prior gives, as $n \to \infty$, a limiting posterior mean , $\sigma^2$, is inconsistent.
  


---

### Noninformative Prior: Reference Prior


The idea of the **reference prior** is to maximize the expected Kullback–Leibler divergence of the posterior distribution relative to the prior. 

This maximizes the expected posterior information about $\theta$ when the prior density is $p(\theta)$; thus, in some sense, p(x) is the "least informative" prior about $\theta$. 

The KL divergence between the prior and posterior distributions is given by

$$KL = \int p(t) \int p(\theta|t) log \frac{p(\theta|t)}{p(\theta)} d\theta dt$$

Here, $t$ is a sufficient statistic for some parameter $\theta$. The inner integral is the KL divergence between the posterior $p(\theta|t)$ and prior $p(\theta)$ distributions and the result is the weighted mean over all values of $t$




Reference priors are often the objective prior of choice in multivariate problems, since other rules (e.g., Jeffreys' rule) may result in priors with problematic behavior.



???

A related idea, reference priors, was introduced by José-Miguel Bernardo.

Here, the idea is to maximize the expected Kullback–Leibler divergence of the posterior distribution relative to the prior. This maximizes the expected posterior information about X when the prior density is p(x); thus, in some sense, p(x) is the "least informative" prior about X.

---

### Noninformative Prior: MAXENT Prior

Another idea is to use the principle of maximum entropy (MAXENT).

The motivation is that the Shannon entropy of a probability distribution measures the amount of information contained in the distribution. 

The larger the entropy, the less information is provided by the distribution. Thus, by maximizing the entropy over a suitable set of probability distributions on $\theta$, one finds the distribution that is least informative in the sense that it contains the least amount of information consistent with the constraints that define the set. 
<br/><br/>

> For example, the maximum entropy prior on a discrete space, given only that the probability is normalized to 1, is the prior that **assigns equal probability to each state**. 

>And in the continuous case, the maximum entropy prior given that the density is normalized with mean zero and unit variance is the **standard normal distribution** $N(0, 1)$.

---



### Weakly Informative Prior


- it expresses partial information about a parameter.

- its main purpose is for regularization, that is, to keep inferences in a reasonable range. It overcomes the possibility for weird values in flat Prior setting.
<br/><br/>

> Example: Setting the prior distribution for the temperature at noon tomorrow

> We may use a normal distribution with mean 50 degrees Fahrenheit and standard deviation 40 degrees, which very loosely constrains the temperature to the range (10 degrees, 90 degrees) with a small chance of being below -30 degrees or above 130 degrees.

---

### Informative Prior

- it is also called as **subjective/substantive** prior.

- it expresses specific, definite information about a variable. 

- it needs our clear understanding of the of-interest parameters. Sometimes, reparameterization is required.

<br/><br/>
> An example is a prior distribution for the temperature at noon tomorrow. 

> A reasonable approach is to make the prior a normal distribution with expected value equal to today's noontime temperature, with variance equal to the day-to-day variance of atmospheric temperature, or a distribution of the temperature for that day of the year.

---

### Conjugate Prior

if the posterior distributions $p(\theta | x)$ are in the same probability distribution family as the prior probability distribution $p(\theta )$, the prior and posterior are then called conjugate distributions, and the prior is called a conjugate prior for the likelihood function $p( x | \theta )$.

- computational convenience

- can be used in both objective and subjective prior settings.
  - $N(0, 10000)$
  - $N(0.4, 0.2)$
  

---

### Example: pharmacokinetics

We illustrate with an example from a model in pharmacokinetics, the study of the absorption, distribution, and elimination of drugs from the body. 

For this particular study, 

- about 20 measurements were available on six young adult males, 

- and a model was fit with 15 parameters per person (which we label $\theta_{kl}$ for person $k$ and parameter $l$), 

- along with two variance parameters, $\sigma_1^2$ and $\sigma^2_2$, indicating the scale of measurement/modeling error. 

The data (concentrations of a compound in blood and exhaled air over time) are only indirectly informative of the individual level parameters, which refer to equilibrium concentrations, volumes, and metabolic rates inside the body.

This is a nice example to use here because different principles for assigning prior distributions are relevant for different parameters in the model, as we
now discuss.

---

### Noninformative Prior Distributions

We first consider the variance parameters $\sigma_1^2$ and $\sigma^2_2$, a noninformative uniform prior distribution works fine. 

The uniform prior distribution here is improper. However, when formally combined with
the data likelihood it yields an acceptable proper posterior distribution.

---

### Highly Informative Prior Distributions

At the other extreme, fairly precise scientific information is available on some of the parameters $\theta_{kl}$ in the model. 

For example, parameter 8 represents the mass of the liver as a fraction of lean body mass; from previous medical studies, the liver is known to be about 3.3% of lean body mass for young adult males, with little variation.



The prior distribution for 

- $\log \theta_{k,8}$ $\sim$ $N(\mu_8, \Sigma_8^2)$. (for persons k = 1,..., 6) 

- $\mu_8 \sim N(\log0.033, \log1.1)$, 

- $\Sigma_8 \sim Inv \chi^2 (2, \log 1.1)$ 

This setup sets the parameters $\theta_{k,8}$ approximately to their prior estimate, 0.033, with relatively small variation allowed between persons.

---
### Moderately Informative Prior Distributions

Finally, some of the parameters $\theta_{kl}$ are not well estimated by the data – thus, they require informative prior distributions – but scientific information on them is limited.

For example, in this particular study, parameter 14 represents the maximum rate of metabolism of a certain compound; the best available estimate of this parameter for healthy humans is 0.042, but this estimate is quite crude and could easily be off by a factor of 10 or 100. The maximum rate of metabolism is not expected to vary greatly between persons, but there is much uncertainty about the numerical value of the parameter. 

This information is encoded in a *hierarchical prior distribution*: 

- $\log \theta_{k, 14} \sim N(\mu_{14}, \Sigma_{14}^2)$ 

- $\mu_{14} \sim N(\log0.042, \log^210)$ 

- $\Sigma_{14} \sim Inv \chi^2 (2, \log^22)$


---

### What Would Happen if Noninformative Prior Distributions Were Used for All the Parameters in this Example?

In our parameterization, noninformative prior distributions on the parameters $\theta_{kl}$ correspond to setting $\Sigma_{.l} = \infty$ for each parameter $l$, thus allowing each person’s parameters to be estimated from that person’s own data. 

If noninformative prior distributions were assigned to all the individual parameters, then the model would fit the data very closely but with
scientifically unreasonable parameters – for example, a person with a liver weighing as value $10$. This sort of difficulty is what motivates a researcher to specify a prior distribution using external information.





---

### Hierarchical: the parameters of the prior



Sometimes we have to set up hyperparameters for the model for 
 - more degree of freedom
 - model the dependence among parameters
 
Hyperparameters themselves may have hyperprior distributions expressing beliefs about their values. 

When things go to hierarchical model, some cares are needed. 

For example, perhaps the most well known theoretical result is that, for variance parameters in a linear regression model, the uniform prior distribution for $\log(\sigma)$ is acceptable when applied to the lowest level variance component but not acceptable for higher level variance components. i.e. $\log \sigma \propto 1$ and $\tau \propto 1$ for proper posterior.

These theoretical results do not give **recommended** models, but rather are useful in **ruling out** certain natural seeming models with poor statistical properties.

---
### Summary: Priors can be created using a number of methods.

- A prior can be determined from past information, such as previous experiments.

- A prior can be elicited from the purely subjective assessment of an experienced expert. 

- An uninformative prior can be created to reflect a balance among outcomes when no information is available. 

- Priors can also be chosen according to some principle, such as symmetry or maximizing entropy given constraints; examples are the Jeffreys prior or Bernardo's reference prior. 

- When a family of conjugate priors exists, choosing a prior from that family simplifies calculation of the posterior distribution.

---
### Five Levels of Priors

- Flat prior;

- Super-vague but proper prior: $N(0, 1e6)$;

- Weakly informative prior, very weak: $N(0, 10)$;

- Generic weakly informative prior: $N(0, 1)$;

- Specific informative prior: $N(0.4, 0.2)$ or whatever. Sometimes this can be expressed as a scaling followed by a generic prior: $\theta = 0.4 + 0.2*z$; $z \sim N(0, 1)$;



---
### Rule of Thumb

- [Documentation of stan software, GitHub](https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations)

```{r echo = FALSE}
knitr::include_url("https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations")
```
---
```{r echo = FALSE}
knitr::include_url("https://yihui.org")
```
---
# Reference

- Andrew Gelman, BDA3

- [Andrew Gelman, Prior distribution](http://www.stat.columbia.edu/~gelman/research/published/p039-_o.pdf)


- [Andrew Gelman, Prior Choice Recommendations](https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations)

- [Surya Tokdar, Choosing a Prior Distribution ](https://www2.stat.duke.edu/courses/Spring13/sta732.01/priors.pdf)

- Jon Wakefield, Bayesian and Frequentist Regression Methods

- [Documentation of stan software, GitHub](https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations)

- [Wikipedia: Prior probability](https://en.wikipedia.org/wiki/Prior_probability#Uninformative_priors)

---

class: inverse, center, middle

# Thanks!
